# **Data Analysis**
***

If you have been following along but stopped, we could load our imported data like so:

Add instructions for loading the data...

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading the data...

</details>
***

<!--
:::{.think_question}
::::{.think_question_header}
Question opportunity
::::
::::{.inner_block}
Pose a question about the data analysis.
<details><summary>Answer</summary>
You can use a collapsible details section to provide an answer
</details>
::::
:::

-->

## Clustering

<!--

Should be able to use `STclust`, but I'm struggling with that. So I grabbed the [cluster assignments from the study](https://github.com/LieberInstitute/spatialDLPFC_SCZ_XENIUM/blob/devel/processed-data/06_cell_type_clustering/banksy_clustering_lambda0.1_res0.7.csv). The [cluster annotations are here in their code](https://github.com/LieberInstitute/spatialDLPFC_SCZ_XENIUM/blob/2c0ec1319137fa1c108ec6ea0c6935b32ad50e1b/code/analysis/06_cell_type_clustering/06_plot_clusters_on_tissue.R#L41-L52).

```{.default}
awk -F, -v OFS=',' 'NR==1 {print "","Cluster","Sample","CoordIndex"} $3 ~ /(Br1139|Br5314|Br5746|Br5973|Br8433|Br8667)/ { gsub(/"/, "", $3); split($3, arr, "_"); print $1, $2, "\""arr[1] "\"", "\""arr[2] "\"" }' data/raw/banksy_clustering_lambda0.1_res0.7.csv > data/wrangled/cluster_assignments.csv
```

Need to take the output of this and join it to the original spatial coordinates. I think that the fourth column is the row number and I need the libname from the original file to join the coordinates to the spatial_meta STlist object

```{r, eval = FALSE}
cluster_assignments <- read.csv(here("data/wrangled/cluster_assignments.csv"))

new_cluster_assignments <- data.table::rbindlist(
  lapply(1:length(spatial_coords), function(x) spatial_coords[[names(spatial_coords)[x]]] %>%
        mutate(CoordIndex = row_number()) %>%
        select(CoordIndex, cell_id) %>%
        right_join(cluster_assignments %>% filter(Sample == substring(names(spatial_coords)[x], 1, 6))) %>%
        rename(libname = cell_id)
)) %>%
  mutate(SampleSuffix = case_when(Sample %in% c("Br8667", "Br5314", "Br8433") ~ "CTRL", 
                                  Sample %in% c("Br5973", "Br1139", "Br5746") ~ "SCZ", 
                                  .default= NA_character_)) %>% 
  unite("SampleName", c(Sample, SampleSuffix), sep = "")
```

-->

```{r}

cluster_assignments <- read.csv(here("data/wrangled/cluster_assignments_withlibname.csv")) %>%
                          rename(annot = Cluster) %>%
                          fill(annot) %>%
                          mutate(annot=case_when(
                                 annot %in% c(6, 8, 1) ~ "Oligo",
                                 annot == 10 ~ "Ambig/Oligo",
                                 annot == 7 ~ "Mic",
                                 annot == 15 ~ "Ambig/In/Endo",
                                 annot == 18 ~ "L5 Ex",
                                 annot == 14 ~ "L6 Ex",
                                 annot == 11 ~ "L4/5 Ex",
                                 annot == 2 ~ "L2/3 Ex",
                                 annot %in% c(17, 16, 4) ~ "Ast",
                                 annot %in% c(5, 13, 3) ~ "Endo",
                                 annot == 12 ~ "CGE",
                                 annot == 9 ~ "MGE",
                                 .default = NA_character_))

for (name in names(scz_spatial_tr@spatial_meta)){

  scz_spatial@spatial_meta[[name]] <- left_join(scz_spatial@spatial_meta[[name]], 
                                                cluster_assignments %>% 
                                                  filter(SampleName == name) %>% 
                                                  select(libname, annot)
                                                )

  scz_spatial_tr@spatial_meta[[name]] <- left_join(scz_spatial_tr@spatial_meta[[name]], 
                                                cluster_assignments %>% 
                                                  filter(SampleName == name) %>% 
                                                  select(libname, annot)
                                                )
  
}
```

### Spatial distribution of expression

:::{.panel-tabset}

#### MBP


```{r}
cellType <- STplot(scz_spatial_tr, plot_meta = "annot", samples = sampleset1)

combined_list <- c(cellType, quiltMBP_sampleset1$plots)

ggpubr::ggarrange(plotlist = combined_list, ncol=2, nrow=2, labels = "AUTO")

```

#### PVALB

```{r}


combined_list2 <- c(cellType, quiltPVALB_sampleset1$plots)

ggpubr::ggarrange(plotlist = combined_list2, ncol=2, nrow=2, labels = "AUTO")

```

:::

### Gene specific summaries of expression

We can also look at summaries of the expression of a few of these genes in these cell types.

```{r}

combine_counts_annot <- function(matannot, matcount, sample_name, gene_name){
    full_join(
        matannot[[sample_name]][,c("libname", "annot")],
        matcount[[sample_name]][gene_name, ] %>% 
            data.frame() %>% 
            rownames_to_column("libname")
    ) %>% `colnames<-`(c("libname", "annot", sample_name))
}

use_combined_data <- function(mat, gene_name, use = "violin", samples = c("Br1139SCZ", "Br5314CTRL")){
  combined_data <- bind_rows(combine_counts_annot(mat@spatial_meta, mat@tr_counts, samples[1], gene_name),
                             combine_counts_annot(mat@spatial_meta, mat@tr_counts, samples[2], gene_name)) %>%
                      pivot_longer(starts_with("Br"),
                                   values_to = "tr_counts",
                                   names_to = "sample") %>%
                      drop_na()

  if (use == "violin"){
    return(
      combined_data %>%
        ggplot(aes(x = sample, 
               y = tr_counts, 
               fill = sample)) + 
        geom_violin() + 
        facet_wrap(~annot, scales = "free_y") + 
        theme_bw() + 
        scale_fill_viridis_d(end = 0.6) + 
        ggtitle(gene_name) +
        theme(legend.position = "bottom")
    )
  } else if (use == "summary"){
    table_length <- length(unique(combined_data$annot)) * 2
    return(
      combined_data %>%
        group_by(annot, sample) %>% 
        summarize(
              min_exp = min(tr_counts), 
              max_exp = max(tr_counts), 
              avg_exp = mean(tr_counts),
              median_exp = median(tr_counts),
              var_exp = var(tr_counts)
        ) %>% 
    print(n=table_length)
    )
  } else {
    message("No other downstream uses supported besides 'violin' and 'summary'")
  }
}

```

:::{.panel-tabset}

#### Log transformation
 
:::{.panel-tabset}

##### MBP

```{r}
use_combined_data(scz_spatial, use = "violin", gene_name = "MBP")
```

```{r}
use_combined_data(scz_spatial, use = "summary", gene_name = "PVALB")
```

##### PVALB

Violin plots of normalized expression by cell type and sample.

```{r}

use_combined_data(scz_spatial, use = "violin", gene_name = "PVALB")

```

Summary tables of normalized expression by cell type and sample.

```{r}
use_combined_data(scz_spatial, use = "summary", gene_name = "PVALB")
```

:::

#### SCTransform

:::{.panel-tabset}

##### MBP

```{r}
use_combined_data(scz_spatial_tr, use = "violin", gene_name = "MBP")
```

```{r}
use_combined_data(scz_spatial_tr, use = "summary", gene_name = "MBP")
```

##### PVALB

```{r}
use_combined_data(scz_spatial_tr, use = "violin", gene_name = "PVALB")
```

```{r}
use_combined_data(scz_spatial_tr, use = "summary", gene_name = "PVALB")
```

:::

:::



***
