# **Data Analysis**
***

If you have been following along but stopped, we could load our imported data like so:

Add instructions for loading the data...

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading the data...

</details>
***

:::{.think_question}
::::{.think_question_header}
Question opportunity
::::
::::{.inner_block}
Pose a question about the data analysis.
<details><summary>Answer</summary>
You can use a collapsible details section to provide an answer
</details>
::::
:::

We will use `SThet` to look at the autocorrelation of the expression of genes throughout the samples.

Moran's I Interpretation:

* Positive: autocorrelation -- similar expression is located near similar expression
* 0: random distribution
* Negative: dissimilar expression is located near each other

Running `SThet` takes 12-15 minutes on our institutional HPC with 2 cores for 2-4 genes across the 6 samples (though only 5 of the 6 run successfully. The sixth fails somewhat silently).

::: {.column-margin}
Why might the final sample be failing? ... I did find this warning message that talks about errors:::
:::

```
Warning messages:
1: In parallel::mclapply(seq(names(x@spatial_meta)), function(i) { :
  1 function calls resulted in an error
2: In parallel::mclapply(seq_along(1:length(unique(as.vector(unlist(combo[[1]]))))),  :
  1 function calls resulted in an error
```

And `get_gene_meta()` takes a couple extra minutes after that. My personal computer can't run this step at all without crashing (though every `spatialGE` step before this it can handle).

::: {.column-margin}
Do you have any recommendations on how to make this combination of `SThet` and `get_gene_meta()` a more feasible step?
:::

Prior to running this step, the data/`STlist` is several hundred MB. Once I've run this step, the data is several hundred GB.

```{r, eval = FALSE}
scz_spatial <- SThet(scz_spatial, genes = c("CX3CR1", "MBP", "FKBP5"))
get_gene_meta(scz_spatial, sthet_only = TRUE)
```

**Possibilities for feasibility**:

* From the documentation, I saw that the k-nearest neighbors approach will be faster, but less accurate, so we could try that.
* We could provide the output and do an analysis task with the output from `get_gene_meta()` -- see thoughts on Z-score below in the "Questions on using this function list."
* We could focus on just two samples instead of 6, but the quilt plot is already using just two samples, so I don't think it makes sense to not use as many as we can here.
* We could subset the data to select a fewer number of cells (randomly drop some cell IDs from the data that we give learners)
* We could compute the statistic for a fewer number of genes (just 2?) ... this didn't seem to speed up the process that much when I tried it

::: {.column-margin}
How do we know if a Moran's I value that is greater than 0 is significantly far from or greater than zero using this function? Does this function compute p-values or help me to compute a z-score?
:::

**Brainstorming on significance**

* If I was trying to compute a z-score from the resulting dataframe ...
  * should I use the `gene_mean` value as the expected value
  * should I use the `gene_stdevs` value as the standard deviation.

If so, I think that's the best possible "analysis" we can have people do if we give them the dataframe ... have them compute the z-score.

So something like ...?

```{r, eval = FALSE}
gene_meta_out <- gene_meta_out %>%
                    summarize(gene_sample_z = (moran_i - gene_mean) / gene_stdevs)
#And do one-tailed tests comparing to difference from 0
```

***
