<!--
Update the file paths as needed throughout
-->

<!--
Some case studies may wish to merge Data Exploration with this Wrangling subsection
-->

<!--
At the end of this section, save incremental data so that users can pause and come back later
-->

# **Data Wrangling**
***

If you have been following along but stopped, we could load our imported data like so:

Add instructions for loading the data...

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading the data...

</details>
***

## Filtering

Note that values used for filtering here are based on results in the data exploration section above. So it and the data exploration section should either be combined in the final case study draft, or we'll add a note about people needing to visit the data exploration step to under the values we're using.

```{r}
scz_spatial <- spatialGE::filter_data(scz_spatial,
                           spot_minreads=50,
                           spot_maxreads=3500,
                           spot_mingenes=10)
```



What do the distributions look like now?

::: {.panel-tabset}

### Table

```{r}
summarize_STlist(scz_spatial) %>% print(width = Inf)
```

### Total Counts

```{r}
distribution_plots(scz_spatial, plot_type='violin', plot_meta='total_counts')
```

### Total Genes

```{r}
distribution_plots(scz_spatial, plot_type='violin', plot_meta='total_genes')
```

:::


## Normalizing

We're using a log transform. Another option is the SCTransform which we'll explain in a click to expand section. We can even compare the two normalization methods (but I'll have to make two separate `STlist`'s for that)

```{r}
scz_spatial <- spatialGE::transform_data(scz_spatial, method='log', cores=2)
```

We can visualize the distribution of expression of a couple of genes (before and after the normalization) using the `distribution_plots()` function.

```{r}
raw_plots <- distribution_plots(scz_spatial, genes = c("CX3CR1", "FKBP5", "SST"), data_type = "raw")

transformed_plots <- distribution_plots(scz_spatial, genes = c("CX3CR1", "FKBP5", "SST"), data_type = "tr")

raw_plots$CX3CR1 + transformed_plots$CX3CR1 + patchwork::plot_layout(guides = "collect")
```

```{r}
raw_plots$FKBP5 + transformed_plots$FKBP5 + plot_layout(guides="collect")
```

```{r}
raw_plots$SST + transformed_plots$SST + plot_layout(guides = "collect")
```

```{r, eval = FALSE}
saveRDS(scz_spatial, here("data/wrangled/scz_spatial.rds"))
```

***