<!--
Update the file paths as needed throughout
-->

<!--
Some case studies may wish to merge Data Exploration with this Wrangling subsection
-->

<!--
At the end of this section, save incremental data so that users can pause and come back later
-->

# **Data Wrangling**
***

If you have been following along but stopped, we could load our imported data like so:

Add instructions for loading the data...

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading the data...

</details>
***

## Filtering

Note that values used for filtering here are based on results in the data exploration section above. So it and the data exploration section should either be combined in the final case study draft, or we'll add a note about people needing to visit the data exploration step to under the values we're using.

```{r}
scz_spatial <- spatialGE::filter_data(scz_spatial,
                           spot_minreads=50,
                           spot_maxreads=3500,
                           spot_mingenes=10)
```

:::{.emphasis_block}
<details><summary>How do you report which cell barcodes were filtered?</summary>
Add info here on how to assess and report which cell barcodes were filtered out. Note that we can compare to the ones that the original study filtered out. 
</details>
:::

What do the distributions look like now?

::: {.panel-tabset}

### Table

```{r}
summarize_STlist(scz_spatial) %>% print(width = Inf)
```

### Total Counts

```{r}
distribution_plots(scz_spatial, plot_type='violin', plot_meta='total_counts')
```

### Total Genes

```{r}
distribution_plots(scz_spatial, plot_type='violin', plot_meta='total_genes')
```

:::


## Normalizing

Transformation with `spatialGE` uses either log-transformation or the `SCTransform` method. 

We'll need a second `STlist` object so we can run two different transformations and compare them.

```{r}
scz_spatial_tr <- rlang::duplicate(scz_spatial)
```

### Log Transformation

Explain log transformation.

```{r}
scz_spatial <- spatialGE::transform_data(scz_spatial, method='log', cores=2)
```

### SCTransform

Explain SCTransform method.

```{r}
scz_spatial_tr <- transform_data(scz_spatial_tr, method = 'sct', cores = 2)
```

### Visualize

We can visualize the library size for each cell within each sample.

```{r}
visualize_lib_size <- function(mat, title_name){
    Matrix::colSums(mat) %>% 
        as.data.frame() %>%
        ggplot(aes(x=`.`)) + geom_histogram(bins = 500) + theme_bw() + ggtitle(title_name)
}

patchwork::wrap_plots(purrr::map2(scz_spatial@counts, names(scz_spatial@counts), visualize_lib_size), nrow = 2) + patchwork::plot_annotation(title = "Raw counts")

wrap_plots(purrr::map2(scz_spatial@tr_counts, names(scz_spatial@tr_counts), visualize_lib_size), nrow = 2) + plot_annotation(title = "Log transformed counts")

wrap_plots(map2(scz_spatial_tr@tr_counts, names(scz_spatial_tr@tr_counts), visualize_lib_size), nrow = 2) + plot_annotation(title = "SCTransform counts)

```

We can visualize the distribution of expression of a few genes (before and after the normalization) using the `distribution_plots()` function.

```{r}
raw_plots <- distribution_plots(scz_spatial, genes = c("CX3CR1", "FKBP5", "PVALB", "MBP"), data_type = "raw")

transformed_log_plots <- distribution_plots(scz_spatial, genes = c("CX3CR1", "FKBP5", "PVALB", "MBP"), data_type = "tr")

transformed_sct_plots <- distribution_plots(scz_spatial_tr, genes = c("CX3CR1", "FKBP5", "PVALB", "MBP"), data_type = "tr")
```

:::{.panel-tabset}

#### CX3CR1

```{r}
raw_plots$CX3CR1 + transformed_log_plots$CX3CR1 + transformed_sct_plots$CX3CR1 + patchwork::plot_layout(guides = "collect", nrow=3)
```

#### FKBP5

```{r}
raw_plots$FKBP5 + transformed_log_plots$FKBP5 + transformed_sct_plots$FKBP5 + patchwork::plot_layout(guides = "collect", nrow=3)
```

#### PVALB

```{r}
raw_plots$PVALB + transformed_log_plots$PVALB + transformed_sct_plots$PVALB + patchwork::plot_layout(guides = "collect", nrow=3)
```

#### MBP

```{r}
raw_plots$MBP + transformed_log_plots$MBP + transformed_sct_plots$MBP + patchwork::plot_layout(guides = "collect", nrow=3)
```

:::

Now we want to clear up some R memory

```{r}
rm(raw_plots, transformed_log_plots, transformed_sct_plots)
gc()
```

```{r, eval = FALSE}
saveRDS(scz_spatial, here("data/wrangled/scz_spatial.rds"))
```

***