# **Data Import**
***

Note that the sample files were fetched with `GEOquery` for `GSE307404`

## 3 random SCZ samples and 3 random NTC samples

1) We set the random seed
2) We read in metadata files, joining them so that we can relate the sample information with the `PrimaryDx` column
3) We select 3 random NTC samples
4) We select 3 random SCZ samples
5) We make a vector of those 6 sample IDs

```{r, eval = FALSE}
set.seed(567)
spatial_samples <- readr::read_csv("/fh/fast/leek_j/user/kisaac/ocs_spatial/spatial_samples.csv")
geo_to_spatial <- read_csv("/fh/fast/leek_j/user/kisaac/ocs_spatial/geo_to_spatial_samples.csv", col_names = FALSE)
spatial_sample_info <- spatial_samples %>%
  filter(Xenium == "YES") %>%
  full_join(geo_to_spatial,
            by = c("Brain Number (ID)" = "X2")) %>%
  select(c("Brain Number (ID)", "Sex", "PrimaryDx", X1))

#3 NTC samples
ntc_samples <- unname(sample(unlist(spatial_sample_info[which(spatial_sample_info$PrimaryDx == "Control"), "X1"]), 3))

#3 SCZ samples
scz_samples <- unname(sample(unlist(spatial_sample_info[which(spatial_sample_info$PrimaryDx == "Schizophrenia"), "X1"]), 3))

samples <- c(scz_samples, ntc_samples)
```

## What are the 3 random samples?

* NTC Samples:
  * 2 females
  * 1 male
* SCZ Samples:
  * all 3 females

## Prepare the data for import into `spatialGE` (with `STlist`)

### Spatial Coordinates

1. We make a list that we can store a dataframe in for each sample
2. We loop through all 6 of the samples .... for each sample ....
3. We read in the corresponding cell boundaries file
4. We group that data by the `cell_id` column since multiple rows are present for each cell ID
5. (with `slice`) We have to append the first row for each cell ID also to the bottom in order to "close the polygon" by going back to the starting vertex
6. We grab coordinates for each cell's polygon and store all coordinates in geometry column
7. (with `st_as_sf()`) we convert the data frame into a simple feature object so that we can work within a spatial context (to compute the cell's centroid) for non-spatial data structures
8. `st_centroid` computes the centroid location for each cell using the `geometry` column from step 6
9. We grab x and y coordinates for the centroids, storing them in their own columns
10. We store the Cell IDs and centroid coordinates as a dataframe for that sample

```{r, eval = FALSE}
spatial_coords <- list()

for (sample in samples){
  df_sample <- read_csv(paste0("/fh/fast/leek_j/user/kisaac/ocs_spatial/GSE307404/",
                               sample,
                               "_",
                               spatial_sample_info[which(spatial_sample_info$X1 == sample), "Brain Number (ID)"],
                               "-cell_boundaries.csv.gz"))

  centroids <- df_sample %>%
    group_by(cell_id) %>%
    slice(1:n(), 1) %>%
    summarize(geometry = sf::st_sfc(st_polygon(list(cbind(vertex_x, vertex_y)))), .groups = "drop") %>%
    st_as_sf() %>%
    mutate(centroid = st_centroid(geometry)) %>% #compute centroid location for each cell
    mutate(cx = st_coordinates(centroid)[,1],
           cy = st_coordinates(centroid)[,2])

  spatial_coords[[sample]] <- centroids %>%
    st_drop_geometry() %>% #so that we can grab just the cell IDs and centroid x,y values
    select(cell_id, cx, cy)
}

```

Note: I know that working with `parquet` or `zarr` files is preferable to this approach. But I don't have `zarr` files, and I'm still working to see if I now have `parquet` files that I can use. We will provide descriptions of these more canonical ways or perhaps switch out for `parquet` files once we can

::: {.column-margin}
**Without access to the canonical files for this task, does this approach seem to be accurate, robust, and a sufficient method in place of those?**
:::

### RNA counts

Note, will show in a click to expand section that `rhdf5::h5dump` can be used to look at the headers of the data, especially because we have to provide a "name" in the reading function and knowing that name requires knowing the headers

1. We make a list that we can store a dataframe in for each sample
2. We loop through all 6 of the samples .... for each sample ....
3. We read in the corresponding cell feature matrix file
4. We build a sparse matrix, using the elements of the h5 file that correspond to each argument (as explained in the data description section above)
5. We expand it to be a dense matrix that can be turned into a dataframe
6. We set the row names to be the feature names from the h5 file
7. We set the column names to be the cell IDs from the h5 file
8. We also move the rownames to the first column and name it "genes" because `spatialGE` looks for gene names in the first column, not as rownames, making sure they are unique. [Got an error here](https://github.com/FridleyLab/spatialGE/blob/849c174f2c1537dc871987dfefb68da91beb96e3/R/STList.R#L415) until I fixed that and it wasn't super obvious to me in the documentation.
9. Finally we store this matrix in our named list for that sample

```{r, eval = FALSE}
rnacounts_input <- list()

for (sample in samples){
  rna_h5data <- rhdf5::h5read(paste0("/fh/fast/leek_j/user/kisaac/ocs_spatial/GSE307404/",
                                     sample,
                                     "_",
                                     spatial_sample_info[which(spatial_sample_info$X1 == sample), "Brain Number (ID)"],
                                     "-cell_feature_matrix.h5"),
                              name = "matrix")

  rnacounts_input[[sample]] <- Matrix::sparseMatrix(i = rna_h5data$indices,
                                                    p = rna_h5data$indptr,
                                                    x = as.integer(rna_h5data$data),
                                                    dims = rna_h5data$shape,
                                                    index1 = FALSE) %>%
                                  as.matrix() %>%
                                  as.data.frame() %>%
                                  `rownames<-`(rna_h5data$features$name) %>%
                                  `colnames<-`(rna_h5data$barcodes) %>%
                                  rownames_to_column("genes")
}
```

### Sample metadata

1. Filter to the 6 random samples
2. Relocate the GEO sample IDs to the first column

```{r, eval = FALSE}
samples_metadata <- spatial_sample_info %>%
  filter(X1 %in% samples) %>%
  relocate(X1)
```

```
# A tibble: 6 × 4
  X1         `Brain Number (ID)` Sex   PrimaryDx
  <chr>      <chr>               <chr> <chr>
1 GSM9223477 Br5314              F     Control
2 GSM9223481 Br8433              M     Control
3 GSM9223479 Br5746              F     Schizophrenia
4 GSM9223484 Br5973              F     Schizophrenia
5 GSM9223483 Br1139              F     Schizophrenia
6 GSM9223485 Br8667              F     Control
```

This is what the sample metadata looks like at this stage. I can bring over more of the metadata columns like Age, `RIN`, and `PMI` depending on the analysis and homework steps.

## Prepare a `spatialGE` project with `STlist`

```{r, eval = FALSE}
scz_spatial <- spatialGE::STlist(rnacounts = rnacounts_input,
                                 spotcoords = spatial_coords,
                                 samples = samples_metadata
                                )
```

Then we'll save the data so people can pick up at the next step after a break if they'd like. In testing the data, I've saved an input list as an `.rds` file before I've input it into `STlist` with no problem. Just want to make sure that the `STlist` object is being saved in an appropriate way that won't cause issues.

::: {.column-margin}
**Would either an `.rda` or `.rds` work for this `STlist` data?**
:::

***
