# **Data Analysis**
***

If you have been following along but stopped, we could load our imported data like so:

Add instructions for loading the data...

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading the data...

</details>
***

<!--
:::{.think_question}
::::{.think_question_header}
Question opportunity
::::
::::{.inner_block}
Pose a question about the data analysis.
<details><summary>Answer</summary>
You can use a collapsible details section to provide an answer
</details>
::::
:::

-->

## Clustering

<!--

Should be able to use `STclust`, but I'm struggling with that. So I grabbed the [cluster assignments from the study](https://github.com/LieberInstitute/spatialDLPFC_SCZ_XENIUM/blob/devel/processed-data/06_cell_type_clustering/banksy_clustering_lambda0.1_res0.7.csv). The [cluster annotations are here in their code](https://github.com/LieberInstitute/spatialDLPFC_SCZ_XENIUM/blob/2c0ec1319137fa1c108ec6ea0c6935b32ad50e1b/code/analysis/06_cell_type_clustering/06_plot_clusters_on_tissue.R#L41-L52).

```{.default}
awk -F, -v OFS=',' 'NR==1 {print "","Cluster","Sample","CoordIndex"} $3 ~ /(Br1139|Br5314|Br5746|Br5973|Br8433|Br8667)/ { gsub(/"/, "", $3); split($3, arr, "_"); print $1, $2, "\""arr[1] "\"", "\""arr[2] "\"" }' data/raw/banksy_clustering_lambda0.1_res0.7.csv > data/wrangled/cluster_assignments.csv
```

Need to take the output of this and join it to the original spatial coordinates. I think that the fourth column is the row number and I need the libname from the original file to join the coordinates to the spatial_meta STlist object

```{r, eval = FALSE}
cluster_assignments <- read.csv(here("data/wrangled/cluster_assignments.csv"))

new_cluster_assignments <- data.table::rbindlist(
  lapply(1:length(spatial_coords), function(x) spatial_coords[[names(spatial_coords)[x]]] %>%
        mutate(CoordIndex = row_number()) %>%
        select(CoordIndex, cell_id) %>%
        right_join(cluster_assignments %>% filter(Sample == substring(names(spatial_coords)[x], 1, 6))) %>%
        rename(libname = cell_id)
)) %>%
  mutate(SampleSuffix = case_when(Sample %in% c("Br8667", "Br5314", "Br8433") ~ "CTRL", 
                                  Sample %in% c("Br5973", "Br1139", "Br5746") ~ "SCZ", 
                                  .default= NA_character_)) %>% 
  unite("SampleName", c(Sample, SampleSuffix), sep = "")
```

-->

```{r}

cluster_assignments <- read.csv(here("data/wrangled/cluster_assignments_withlibname.csv"))

for (name in names(scz_spatial_tr@spatial_meta)){
  scz_spatial_tr@spatial_meta[[name]] <- left_join(scz_spatial_tr@spatial_meta[[name]], 
                                                cluster_assignments %>% 
                                                  filter(SampleName == name) %>% 
                                                  select(libname, Cluster)
                                                )  %>%
                                          rename(annot = Cluster) %>%
                                          fill(annot) %>%
                                          mutate(annot=case_when(
                                            annot %in% c(6, 8, 1) ~ "Oligo",
                                            annot == 10 ~ "Ambig/Oligo",
                                            annot == 7 ~ "Mic",
                                            annot == 15 ~ "Ambig/In/Endo",
                                            annot == 18 ~ "L5 Ex",
                                            annot == 14 ~ "L6 Ex",
                                            annot == 11 ~ "L4/5 Ex",
                                            annot == 2 ~ "L2/3 Ex",
                                            annot %in% c(17, 16, 4) ~ "Ast",
                                            annot %in% c(5, 13, 3) ~ "Endo",
                                            annot == 12 ~ "CGE",
                                            annot == 9 ~ "MGE",
                                            .default = NA_character_))
                                                

}
```

### Observed cell types

We can look at the proportion of cells in each sample that are each cell type.

```
num_cell_types <- function(mat, colname){
  return(mat %>% 
    group_by(annot) %>% summarize("{colname}" := n())
  )

}

bind_rows(map2(scz_spatial@spatial_meta, 
               names(scz_spatial@spatial_meta), 
               num_cell_types)) %>% 
    pivot_longer(starts_with("Br"), 
                names_to = "sample", 
                values_to = "count") %>% 
    drop_na() %>% 
    group_by(sample) %>% 
    mutate(total_cells = sum(count), 
           prop = count / total_cells * 100, 
           sample = factor(sample, 
                           levels = c("Br5314CTRL", "Br8433CTRL", "Br8667CTRL", "Br1139SCZ", "Br5746SCZ", "Br5973SCZ"))) %>% 
    ggplot(aes(x = sample, 
               y = prop, 
               fill = annot)) + 
    geom_bar(stat = "identity") + 
    theme_bw() + 
    scale_fill_viridis_d() +
    ylab("Proportion of cells")
```

### Gene specific summaries of expression

We can also look at summaries of the expression of a few of these genes in these cell types. 

:::{panel.tabet}

#### PVALB

```{r}
combine_counts_annot <- function(matannot, matcount, sample_name, gene_name){
  full_join(
    matannot[[sample_name]][,c("libname", "annot")],
    matcount[[sample_name]][gene_name, ] %>% 
      data.frame() %>% 
      rownames_to_column("libname")
  ) %>% `colnames<-`(c("libname", sample_name, "annot"))
}

bind_rows(combine_counts_annot(scz_spatial@spatial_meta, scz_spatial@tr_counts, "Br1139SCZ", "PVALB"), 
          combine_counts_annot(scz_spatial@spatial_meta, scz_spatial@tr_counts, "Br5314CTRL", "PVALB")) %>% 
    pivot_longer(starts_with("Br"), 
                 values_to = "tr_counts", 
                 names_to = "sample") %>% 
    drop_na() %>% 
    ggplot(aes(x = sample, 
               y = tr_counts, 
               fill = sample)) + 
    geom_violin() + 
    facet_wrap(~annot) + 
    theme_bw() + 
    scale_fill_viridis_d(end = 0.6) + 
    ggtitle("PVALB")
```

```{r}
bind_rows(combine_counts_annot(scz_spatial@spatial_meta, scz_spatial@tr_counts, "Br1139SCZ", "PVALB"), 
          combine_counts_annot(scz_spatial@spatial_meta, scz_spatial@tr_counts, "Br5314CTRL", "PVALB")) %>% 
    pivot_longer(starts_with("Br"), 
                 values_to = "tr_counts", 
                 names_to = "sample") %>% 
    drop_na() %>% 
    group_by(annot, sample) %>% 
    summarize(min_exp = min(tr_counts), 
              max_exp = max(tr_counts), 
              avg_exp = mean(tr_counts),
              median_exp = median(tr_counts)) %>% 
    print(n=24)
```

#### MBP

```{r}

bind_rows(combine_counts_annot(scz_spatial@spatial_meta, scz_spatial@tr_counts, "Br1139SCZ", "MBP"), 
          combine_counts_annot(scz_spatial@spatial_meta, scz_spatial@tr_counts, "Br5314CTRL", "MBP")) %>% 
    pivot_longer(starts_with("Br"), 
                 values_to = "tr_counts", 
                 names_to = "sample") %>% 
    drop_na() %>% 
    ggplot(aes(x = sample, 
               y = tr_counts, 
               fill = sample)) + 
    geom_violin() + 
    facet_wrap(~annot) + 
    theme_bw() + 
    scale_fill_viridis_d(end = 0.6) + 
    ggtitle("MBP")
```

```{r}
bind_rows(combine_counts_annot(scz_spatial@spatial_meta, scz_spatial@tr_counts, "Br1139SCZ", "MBP"), 
          combine_counts_annot(scz_spatial@spatial_meta, scz_spatial@tr_counts, "Br5314CTRL", "MBP")) %>% 
    pivot_longer(starts_with("Br"), 
                 values_to = "tr_counts", 
                 names_to = "sample") %>% 
    drop_na() %>% 
    group_by(annot, sample) %>% 
    summarize(min_exp = min(tr_counts), 
              max_exp = max(tr_counts), 
              avg_exp = mean(tr_counts),
              median_exp = median(tr_counts)) %>% 
    print(n=24)
```

::::

### Spatial distribution of expression

:::{.panel-tabset}

#### PVALB

```{r}
cellType <- STplot(scz_spatial_tr, plot_meta = "annot", samples = sampleset1)

combined_list <- c(cellType, quiltPVALB_sampleset1$plots)

ggpubr::ggarrange(plotlist = combined_list, ncol=2, nrow=2, labels = "AUTO")

```

#### MBP


```{r}
combined_list2 <- c(cellType, quiltMVP_sampleset1$plots)

ggpubr::ggarrange(plotlist = combined_list2, ncol=2, nrow=2, labels = "AUTO")

```

:::

***
