# **Data Import**
***

Since we are working with Xenium spatial transcriptomics data for multiple samples, we need to provide `spatialGE` with 3 types of data.

1. Sample Metadata (primarily the sample IDs)
2. Spatial Coordinates
3. The mRNA counts


`spatialGE` provides the user some flexibility in how they import their data. Users can either provide file paths or they can provide named lists with each list containing data frames for 

Provide instructions on how to get the `spatial_samples.csv` and `geo_to_spatial_samples.csv` files.

## 3 random SCZ samples and 3 random NTC samples

1) We set the random seed
2) We read in metadata files, joining them so that we can relate the sample information with the `PrimaryDx` column
3) We select 3 random NTC samples
4) We select 3 random SCZ samples
5) We make a vector of those 6 sample IDs

```{r message = FALSE, warning = FALSE}
set.seed(567)
spatial_samples <- readr::read_csv(here::here("data/raw/spatial_samples.csv"))
geo_to_spatial <- read_csv(here("data/raw/geo_to_spatial_samples.csv"), col_names = FALSE)
spatial_sample_info <- spatial_samples %>%
  dplyr::filter(Xenium == "YES") %>%
  full_join(geo_to_spatial,
            by = c("Brain Number (ID)" = "X2")) %>%
  select(c("Brain Number (ID)", "Sex", "PrimaryDx", X1))

#3 NTC samples
ntc_samples <- unname(sample(unlist(spatial_sample_info[which(spatial_sample_info$PrimaryDx == "Control"), "X1"]), 3))

#3 SCZ samples
scz_samples <- unname(sample(unlist(spatial_sample_info[which(spatial_sample_info$PrimaryDx == "Schizophrenia"), "X1"]), 3))

samples <- c(scz_samples, ntc_samples)

samples
```

## What are the 3 samples subset from the larger cohort for each condition?

* NTC Samples:
  * 2 females
  * 1 male
* SCZ Samples:
  * all 3 females

```{r}
spatial_sample_info[spatial_sample_info$X1 %in% samples,]
```

## Fetch GEO data for these samples

This `lapply` will loop through the samples vector and fetch the following files for each sample:

* `-cell_boundaries.csv.gz` for computing spatial coordinates 
* `parquet.gz` for spatial coordinates
* `-cell_feature_matrix.h5` for RNA expression

```{r}
lapply(1:length(samples), 
       function(x) 
       GEOquery::getGEOSuppFiles(GEO = samples[x], 
                      filter_regex = "\\-cell_boundaries|\\-cell_feature_matrix|\\parquet", 
                      baseDir = here("data/raw")
                      )
      )
```

## Prepare the data for import into `spatialGE` (with `STlist`)

### Sample metadata

1. Rename the column with a space in its name to have no spaces
2. Filter to include only the subset 6 samples
3. Combine the Brain Number ID with the Primary Dx to create an ID that provides more information about each sample
4. Relocate this new ID to the first column
5. Arrange so the rows are in the same order as the `samples` vector.

<!--
I can bring over more of the metadata columns like Age, `RIN`, and `PMI` depending on the analysis and homework steps.
-->


```{r}
lookup <- c(sampleID = "X1", brainNumID = "Brain Number (ID)")
samples_metadata <- spatial_sample_info %>% 
  dplyr::rename(., all_of(lookup)) %>%
  filter(sampleID %in% samples) %>%
  mutate(PrimaryDx = case_when(PrimaryDx == "Control" ~ "CTRL", 
                               PrimaryDx == "Schizophrenia" ~ "SCZ")
        ) %>% 
  tidyr::unite("DxID", 
               brainNumID, PrimaryDx, 
               sep = "", 
               remove = FALSE) %>%
  relocate(DxID) %>%
  arrange(match(sampleID, samples))

samples_metadata  
```

Let's get rid of the other data frames so we don't accidentally use them going forward and to reclaim some memory.

```{r}
rm(spatial_samples, geo_to_spatial, spatial_sample_info, ntc_samples, scz_samples, samples, lookup)
```

### Spatial Coordinates

```{r}
decompress_read_parquet <- function(samples_metadata, sample){
  #Build file paths
  gz_file <- here(paste0("data/raw/",
                         sample,
                         "/",
                         sample,
                         "_",
                         samples_metadata[which(samples_metadata$sampleID == sample), "brainNumID"],
                         "-cells.parquet.gz"))
  
  decompressed_file <- str_replace(gz_file, ".gz", "")

  #Decompress
  R.utils::gunzip(gz_file, destname = decompressed_file, remove = FALSE, skip = TRUE)

  #Read in
  df <- arrow::read_parquet(decompressed_file)

  #Remove the decompressed file because we no longer need it
  file.remove(decompressed_file)
  
  #Select columns of interest
  centroids <- df %>%
    select(cell_id, x_centroid, y_centroid)

  #reclaim some R resources
  rm(df, gz_file, decompressed_file)
  gc()

  #Return
  return(centroids)

}

spatial_coords <- lapply(1:nrow(samples_metadata), function(x) decompress_read_parquet(samples_metadata, samples_metadata$sampleID[x]))
names(spatial_coords) <- samples_metadata$DxID
```

<details><summary>What if you don't have the canonical parquet files?</summary>

1. We make a list that we can store a dataframe in for each sample
2. We loop through all 6 of the samples .... for each sample ....
3. We read in the corresponding cell boundaries file
4. We group that data by the `cell_id` column since multiple rows are present for each cell ID
5. (with `slice`) We have to append the first row for each cell ID also to the bottom in order to "close the polygon" by going back to the starting vertex
6. We grab coordinates for each cell's polygon and store all coordinates in geometry column
7. (with `st_as_sf()`) we convert the data frame into a simple feature object so that we can work within a spatial context (to compute the cell's centroid) for non-spatial data structures
8. `st_centroid` computes the centroid location for each cell using the `geometry` column from step 6
9. We grab x and y coordinates for the centroids, storing them in their own columns
10. We store the Cell IDs and centroid coordinates as a dataframe for that sample


```{r spatial_coords_compute}

testsample <- samples_metadata$sampleID[1]

df_test_sample <- read_csv(here(paste0("data/raw/",
                                 testsample,
                                 "/",
                                 testsample,
                                 "_",
                                 samples_metadata[which(samples_metadata$sampleID == testsample), "brainNumID"],
                               "-cell_boundaries.csv.gz")))

centroids <- df_test_sample %>%
    group_by(cell_id) %>%
    slice(1:n(), 1) %>%
    summarize(geometry = sf::st_sfc(st_polygon(list(cbind(vertex_x, vertex_y)))), .groups = "drop") %>%
    st_as_sf() %>%
    mutate(centroid = st_centroid(geometry)) %>% #compute centroid location for each cell
    mutate(cx = st_coordinates(centroid)[,1],
           cy = st_coordinates(centroid)[,2]) %>%
    st_drop_geometry() %>% #so that we can grab just the cell IDs and centroid x,y values
    select(cell_id, cx, cy)
```

This method is accurate if compared to the parquet files.

```{r}
right_join(spatial_coords[[unlist(samples_metadata[which(samples_metadata$sampleID == testsample), "DxID"])]],
           centroids,
           by = "cell_id") %>%
    rename(x_dataset_parquet = x_centroid, 
           y_dataset_parquet = y_centroid, 
           x_dataset_compute = cx, 
           y_dataset_compute = cy) %>% 
    pivot_longer(cols = - cell_id, 
                 names_to = c("variable", "dataset"), 
                 names_pattern = "(x|y)_(.*)") %>% 
    pivot_wider(names_from = dataset, 
                values_from = value) %>% 
    ggplot(aes(x = dataset_parquet, 
               y = dataset_compute)) + 
      geom_point(alpha = 0.2) + 
      facet_wrap(~variable) + 
      ggpubr::stat_regline_equation(aes(label =  paste(..eq.label.., ..rr.label.., sep = "*`,`~")), 
                                    label.x = 3, 
                                    label.y = 7500)

```

</details>

### RNA counts

Note, will show in a click to expand section that `rhdf5::h5dump` can be used to look at the headers of the data, especially because we have to provide a "name" in the reading function and knowing that name requires knowing the headers


2. We create a function to use for all 6 of the samples .... for each sample ....
3. We read in the corresponding cell feature matrix file
4. We build a sparse matrix, using the elements of the h5 file that correspond to each argument (as explained in the data description section above)
5. We expand it to be a dense matrix that can be turned into a dataframe
6. We set the row names to be the feature names from the h5 file
7. We set the column names to be the cell IDs from the h5 file
8. We also move the rownames to the first column and name it "genes" because [`spatialGE` looks for gene names in the first column, not as rownames, making sure they are unique.](https://github.com/FridleyLab/spatialGE/blob/849c174f2c1537dc871987dfefb68da91beb96e3/R/STList.R#L415).
9. Finally we store this matrix in our named list for that sample

:::{.column-margin}
Turn this into an `lapply` statement instead of a `for` loop, utilizing a function for the main steps.
:::

```{r}
read_rna <- function(samples_metadata, sample){

  h5file <- here(paste0("data/raw/",
                         sample,
                         "/",
                         sample,
                         "_",
                         samples_metadata[which(samples_metadata$sampleID == sample), "brainNumID"],
                        "-cell_feature_matrix.h5"))
                              

  rna_h5data <- rhdf5::h5read(h5file, name = "matrix")

  rna_mat <- Matrix::sparseMatrix(i = rna_h5data$indices,
                       p = rna_h5data$indptr,
                       x = as.integer(rna_h5data$data),
                       dims = rna_h5data$shape,
                       index1 = FALSE) %>%
      as.matrix() %>%
      as.data.frame() %>%
      `rownames<-`(rna_h5data$features$name) %>%
      `colnames<-`(rna_h5data$barcodes) %>%
      rownames_to_column("genes")

  rm(rna_h5data, h5file)
  gc()

  return(rna_mat)
}

rnacounts_input <- lapply(1:nrow(samples_metadata), function(x) read_rna(samples_metadata, samples_metadata$sampleID[x]))
names(rnacounts_input) <- samples_metadata$DxID

```

## Prepare a `spatialGE` project with `STlist`

```{r}
scz_spatial <- spatialGE::STlist(rnacounts = rnacounts_input,
                                 spotcoords = spatial_coords,
                                 samples = samples_metadata
                                )

rm(rnacounts_input, spatial_coords)
gc()                               
```

Then we'll save the data so people can pick up at the next step after a break if they'd like. In testing the data, I've saved an input list as an `.rds` file before I've input it into `STlist` with no problem. Just want to make sure that the `STlist` object is being saved in an appropriate way that won't cause issues.

```{r, eval = FALSE}
saveRDS(scz_spatial, here("data/imported/scz_spatial.rds"))
```

***
