# **Data Analysis**
***

If you have been following along but stopped, we could load our imported data like so:

Add instructions for loading the data...

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading the data...

</details>
***

<!--
:::{.think_question}
::::{.think_question_header}
Question opportunity
::::
::::{.inner_block}
Pose a question about the data analysis.
<details><summary>Answer</summary>
You can use a collapsible details section to provide an answer
</details>
::::
:::

-->

## Clustering

Should be able to use `STclust`, but I'm struggling with that. So I grabbed the [cluster assignments from the study](https://github.com/LieberInstitute/spatialDLPFC_SCZ_XENIUM/blob/devel/processed-data/06_cell_type_clustering/banksy_clustering_lambda0.1_res0.7.csv). The [cluster annotations are here in their code](https://github.com/LieberInstitute/spatialDLPFC_SCZ_XENIUM/blob/2c0ec1319137fa1c108ec6ea0c6935b32ad50e1b/code/analysis/06_cell_type_clustering/06_plot_clusters_on_tissue.R#L41-L52).

<!--
```{.default}
awk -F, -v OFS=',' 'NR==1 {print "","Cluster","Sample","CoordIndex"} $3 ~ /(Br1139|Br5314|Br5746|Br5973|Br8433|Br8667)/ { gsub(/"/, "", $3); split($3, arr, "_"); print $1, $2, "\""arr[1] "\"", "\""arr[2] "\"" }' data/raw/banksy_clustering_lambda0.1_res0.7.csv > data/wrangled/cluster_assignments.csv
```

Need to take the output of this and join it to the original spatial coordinates. I think that the fourth column is the row number and I need the libname from the original file to join the coordinates to the spatial_meta STlist object

```{r, eval = FALSE}
cluster_assignments <- read.csv(here("data/wrangled/cluster_assignments.csv"))

new_cluster_assignments <- data.table::rbindlist(
  lapply(1:length(spatial_coords), function(x) spatial_coords[[names(spatial_coords)[x]]] %>%
        mutate(CoordIndex = row_number()) %>%
        select(CoordIndex, cell_id) %>%
        right_join(cluster_assignments %>% filter(Sample == substring(names(spatial_coords)[x], 1, 6))) %>%
        rename(libname = cell_id)
)) %>%
  mutate(SampleSuffix = case_when(Sample %in% c("Br8667", "Br5314", "Br8433") ~ "CTRL", 
                                  Sample %in% c("Br5973", "Br1139", "Br5746") ~ "SCZ", 
                                  .default= NA_character_)) %>% 
  unite("SampleName", c(Sample, SampleSuffix), sep = "")

for (name in names(scz_spatial@spatial_meta)){
  scz_spatial@spatial_meta[[name]] <- left_join(scz_spatial@spatial_meta[[name]], 
                                                new_cluster_assignments %>% 
                                                  filter(SampleName == name) %>% 
                                                  select(libname, Cluster)
                                                )  %>%
                                          rename(annot = Cluster) %>%
                                          fill(annot) %>%
                                          mutate(annot=case_when(
                                            annot %in% c(6, 8, 1) ~ "Oligo",
                                            annot == 10 ~ "Ambig/Oligo",
                                            annot == 7 ~ "Mic",
                                            annot == 15 ~ "Ambig/In/Endo",
                                            annot == 18 ~ "L5 Ex",
                                            annot == 14 ~ "L6 Ex",
                                            annot == 11 ~ "L4/5 Ex",
                                            annot == 2 ~ "L2/3 Ex",
                                            annot %in% c(17, 16, 4) ~ "Ast",
                                            annot %in% c(5, 13, 3) ~ "Endo",
                                            annot == 12 ~ "CGE",
                                            annot == 9 ~ "MGE",
                                            .default ~ NA_character_))
                                                

}
```

-->

We will use `SThet` to look at the autocorrelation of the expression of genes throughout the samples.

Moran's I Interpretation:

* Positive: autocorrelation -- similar expression is located near similar expression
* 0: random distribution
* Negative: dissimilar expression is located near each other

Running `SThet` takes 12-15 minutes on our institutional HPC with 2 cores for 2-4 genes across the 6 samples (though only 5 of the 6 run successfully. The sixth fails somewhat silently).

::: {.column-margin}
**Why might the final sample be failing? ... I did find this warning message that talks about errors**
:::

```
Warning messages:
1: In parallel::mclapply(seq(names(x@spatial_meta)), function(i) { :
  1 function calls resulted in an error
2: In parallel::mclapply(seq_along(1:length(unique(as.vector(unlist(combo[[1]]))))),  :
  1 function calls resulted in an error
```

And `get_gene_meta()` takes a couple extra minutes after that. My personal computer can't run this step at all without crashing (though every `spatialGE` step before this it can handle).


Prior to running this step, the data/`STlist` is several hundred MB. Once I've run this step, the data is several hundred GB.

```{r, eval = FALSE}
scz_spatial <- SThet(scz_spatial, genes = c("CX3CR1", "MBP", "FKBP5"))
get_gene_meta(scz_spatial, sthet_only = TRUE)
```

**Possibilities for feasibility**:

* From the documentation, I saw that the k-nearest neighbors approach will be faster, but less accurate, so we could try that.
* We could provide the output and do an analysis task with the output from `get_gene_meta()` -- see thoughts on Z-score below in the "Questions on using this function list."
* We could focus on just two samples instead of 6, but the quilt plot is already using just two samples, so I don't think it makes sense to not use as many as we can here.
* We could subset the data to select a fewer number of cells (randomly drop some cell IDs from the data that we give learners)
* We could compute the statistic for a fewer number of genes (just 2?) ... this didn't seem to speed up the process that much when I tried it
* Focus on a specific subregion of the tissue only? Would I need to have or prepare a separate STlist object for this?

::: {.column-margin}
**Do you have any recommendations on how to make this combination of `SThet` and `get_gene_meta()` a more feasible step?**
:::

**Brainstorming on significance**

* If I was trying to compute a z-score from the resulting dataframe ...
  * should I use the `gene_mean` value as the expected value
  * should I use the `gene_stdevs` value as the standard deviation.

::: {.column-margin}
**How do we know if a Moran's I value that is greater than 0 is significantly far from or greater than zero using this function? Does this function compute p-values or help me to compute a z-score?**
:::


If so, I think that's the best possible "analysis" we can have people do if we give them the dataframe ... have them compute the z-score.

So something like ...?

```{r, eval = FALSE}
gene_meta_out <- gene_meta_out %>%
                    summarize(gene_sample_z = (moran_i - gene_mean) / gene_stdevs)
#And do one-tailed tests comparing to difference from 0
```

***
