# **Data Exploration**
***

Add instructions for loading the data...

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading the data...

</details>
***

```{r}
spatialGE::summarize_STlist(scz_spatial) %>% print(width = Inf)
```

This is a 6 X 9 dataframe that provides

  1. `sample_name`: GEO IDs we provided in sample metadata
  2. `spotscells`: how many spots or cells that sample has
  3. `genes`: how many genes that sample has
  4. `min_counts_per_spotcell`: minimum number of counts per spot or cell for that sample
  5. `mean_counts_per_spotcell`: average number of counts per spot or cell for that sample
  6. `max_counts_per_spotcell`: maximum number of counts per spot or cell for that sample
  7. `min_genes_per_spotcell`: minimum number of genes (with non-zero counts?) per spot or cell for that sample
  8. `mean_genes_per_spotcell`: average number of genes (with non-zero counts?) per spot or cell for that sample
  9. `max_genes_per_spotcell`: maximum number of genes (with non-zero counts?) per spot or cell for that sample

Violin plot of counts per spot/cell for each sample

```{r}
spatialGE::distribution_plots(scz_spatial, plot_type='violin', plot_meta='total_counts')
```

Violin plot of genes per spot/cell for each sample

```{r}
distribution_plots(scz_spatial, plot_type='violin', plot_meta='total_genes')
```

What do we observe:

* There aren't any samples that have a huge number of counts or genes higher than other samples, so I don't think we'll need to use a spot_max* argument within the `filter_data` function
* Every sample has at least 0 or 2 counts as a minimum (I'm surprised that it reports 0 and not the lowest non-zero value)
* Every sample has at least 0 or 2 genes as a minimum as well.
* Mean counts is around 600-900, with maxes in the thousands.
* Going to set a minimum of 500 reads/counts for each spot/cell since that's a little below the mean.
* Mean genes is around 80-110, with maxes around 200.
* Going to set a minimum of 50 genes for each spot/cell since that's a little below the mean.
* Some samples have a lot more spots/cells than others .... I should probably make a table after filtering to see if there are a similar number of spots/cells after filtering. But I also figure that means the tissue is different sizes?

Won't need to save data at the end of this step because we haven't changed anything.
