# **Data Visualization**
***


If you have been following along but stopped, we could load our imported data like so:

Add instructions for loading the data...

***
<details> <summary> If you skipped the data import section click here. </summary>

Add instructions for loading the data...

</details>
***

We are going to look at the expression of the genes FKBP5, CX3CR1, NTRK2 (thought question?), MBP, PVALB, and BDNF (thought question?). For each gene we are going to consider both the amount of its expression as well as where it is expressed (spatially) in the tissue samples. For each gene, we can visually compare its spatial expression in a sample from a neurotypical participant to a sample from a participant with Schizophrenia.

To visually explore the spatial distribution of expression of these genes, we are going to plot Quilt plots.

We're going to make quilt plots showing the expression of these genes for each spot/cell in two samples (one NTC and one SCZ).

Note that quilt plots don't require histology image (image of the tissue), because the spatial coordinates data that we input provides x and y coordinates for the center of each cell/spot. Therefore, we can place the location of these cells without a tissue image.

Also note that this is Xenium data which means there are a lot of cells! So we're going to use the `ptsize` argument to set the point size of the spots or cells to be really small. If we make it too large, it's dense/cramped/hard to view the quilt plot

I tested several color palettes and settled on `PuBuGn`.... more explanation below.

We have `ggpubr` combine the common legends and display them on the bottom. It also is plotting the quilt plots in a 2 X 2 grid

## Differential Expression Only

### CX3CR1

::: {.panel-tabset}

#### Set of samples 1

```{r diff_exp1_quilt}
quilt26CX3CR1 <- spatialGE::STplot(scz_spatial_tr,
                  samples = samples_metadata$DxID[c(2,6)],
                  genes = c("CX3CR1"),
                  color_pal='PuBuGn',
                  ptsize=0.05)

quiltArrange <- ggpubr::ggarrange(plotlist=quilt26CX3CR1,
                  nrow=1,
                  ncol=2,
                  legend='bottom',
                  common.legend = TRUE)

quiltCX3CR1
```

#### Set of samples 2

```{r}
displayQuilt <- function(stlist, sampleNames, gene){

    quiltPlots <- spatialGE::STplot(stlist,
                                    samples = sampleNames,
                                    genes = gene,
                                    color_pal = 'PuBuGn',
                                    ptsize=0.05
                                   )

    quiltArrange <- ggpubr::ggarrange(plotlist=quiltPlots,
                                      nrow = 1,
                                      ncol = 2,
                                      legend = 'bottom',
                                      common.legend = TRUE)

    return(quiltArrange)

}
```
:::



```{r}
sampleset2 <- samples_metadata$DxID[c(1,4)]
quiltCX3CR1_sampleset2 <- displayQuilt(scz_spatial_tr, 
                                       sampleset2,
                                       "CX3CR1")

quiltCX3CR1_sampleset2

```

:::

### FKBP5

::: {.panel-tabset}

#### Set of samples 1

```{r diff_exp2_quilt}

sampleset1 <- samples_metadata$DxID[c(2,6)]
quiltFKBP5_sampleset1 <- displayQuilt(scz_spatial_tr, 
                                       sampleset1,
                                       "FKBP5")

quiltFKBP5_sampleset1

```

#### Set of samples 2

```{r}
quiltFKBP5_sampleset2 <- displayQuilt(scz_spatial_tr, 
                                       sampleset2,
                                       "FKBP5")

quiltFKBP5_sampleset2

```

:::

## Differential Spatial Distribution

### MBP

::: {.panel-tabset}

#### Set of samples 1

```{r diff_exp2_quilt}

quiltMBP_sampleset1 <- displayQuilt(scz_spatial_tr, 
                                       sampleset1,
                                       "MBP")

quiltMBP_sampleset1

```

#### Set of samples 2

```{r}
quiltMBP_sampleset2 <- displayQuilt(scz_spatial_tr, 
                                       sampleset2,
                                       "MBP")

quiltMBP_sampleset2

```

:::

### PVALB

::: {.panel-tabset}

#### Set of samples 1

```{r diff_exp2_quilt}

quiltPVALB_sampleset1 <- displayQuilt(scz_spatial_tr, 
                                       sampleset1,
                                       "PVALB")

quiltPVALB_sampleset1

```

#### Set of samples 2

```{r}
quiltPVALB_sampleset2 <- displayQuilt(scz_spatial_tr, 
                                       sampleset2,
                                       "PVALB")

quiltPVALB_sampleset2
```

:::




FKBP5 is annotated in the Figure 2a Volcano Plot as DEGs. I went to [Supplementary Table 4 from the paper](https://www.biorxiv.org/content/biorxiv/early/2026/02/17/2026.02.16.706214/DC5/embed/media-5.xlsx?download=true) and found the gene with the lowest / most negative log fold change: CX3CR1

**Color Palettes**

I tested several color palettes and settled on `PuBuGn` for several reasons. (Note: this could be a cool side quest to include in the case study)

1. Easier to distinguish changes in expression (e.g., up vs down regulated). `BuRd` and `Spectral` were best for this otherwise.
2. The lighter color was no longer in the center (as if the log expression values were "centered") and instead is on the lower level of the spectrum. `Spectral` was the only other one not centered on white, but it was still centered on a lighter color.
3. Looking at the result with a color vision deficiency simulator (`cvg_grid()` from `colorblindr`) the general conclusions are still possible in my opinion. This wasn't the case with `PrGn`, `BuRd`, or `Spectral` when I tested them.

```{r}
colorblindr::cvd_grid(plot = quiltArrange)
```

<!--
Look at these genes to try to see differences in spatial distribution not just up/down: MBP, GFAP, NTRK2, MT1E, PVALB, BDNF
-->

Let's clear some R memory again.

```{r}
rm(quilt26CX3CR1, quilt26FKBP5, quilt26MBP, quilt26PVALB, quiltArrange)
gc()
```

***
